Glucose ingestion before and after resistance training does not augment ribosome biogenesis in healthy moderately trained young adults

### 2023-03-15 ribose-paper.RMD
updated discussion

Responses to systematic resistance training (RT) vary widely between individuals, with as many as 10-15% showing impaired muscle growth in response to standardized training interventions (ALVAREZ18;MANN14;THALACKER13). Such inability to benefit from RT may be explained by genetic and/or epigenetic disposition, affecting the internal physiological milieu in terms of ability to mount an anabolic response to RT [@figueiredo_genetic_2021;@thalacker-mercer_cluster_2013].<!-- Check these refs, do they correspond to genet epigen and internal mileu? Kristian: The way I read it, it looks like it corresponds --> Although RT program modifications such as altered exercise volume or intensity may induce more favorable adaptations in some individuals, such benefits are not uniform between individuals [@hammarstrom_benefits_2020;@krieger_single_2009;@schoenfeld_dose-response_2017].<!-- Add ref for more rt program modifications. Kristian: Added krieger et al. 2009 and schoenfeld et al. 2017 --> For instance, although changes in training volume generally induces greater benefits of RT, not all individuals benefit from such modifications [@hammarstrom_benefits_2020]. Therefore, other means than modification of RT variables seems necessary to optimize individual responses.<!-- The above sentence might be a bit contradictory? Im not sure, are you thinking about "other means than modification of RT" being contradictory of "modifications such as altered exercise volume.." sentence? I think there is a point to establishing that yes, increasing exercise volume is a potent way to increase the RT stimulus in general. However, in cases where this alone fails to elicit true responses, other means are necessary. Or am I missinterpreting this? -->

@@

To our knowledge, no previous study has investigated the effects of glucose compared to placebo ingestion with RT on ribosome biogenesis in human skeletal muscle, making our hypotheses exploratory and comparisons not straight forwards. Mariappan and colleagues (2011) investigated the effect of high glucose treatment on ribosome biogenesis in glomerular epithelial cells (GECs) from diabetic (type 1 and 2) mice, observing a UBF-dependent augmentation of ribosome biogenesis. In a study from 2015, Tanaka et al. (2015) observed that mild glucose starvation reduced rRNA transcription in human breast cancer cells. These findings do imply an important role for glucose in the initiation of rDNA transcription, the rate-limiting step of ribosome biogenesis [@mariappan_ribosomal_2011; @tanaka_mild_2015]. In the present study, we did not find such indications of glucose having an effect as these previous studies. Notably, kidney cells from mice and human breast cancer cells are likely to have a higher physiological rate compared to human skeletal muscle cells, thus, our two-week intervention may not be directly comparable to these findings, as was expected. Nevertheless, we did observe robust increases in both total RNA and 47S pre-rRNA and all four mature rRNAs (18S, 28S, 5.8S, 5S) after five RT sessions with both treatments. This means that the time frame and training volume chosen for this design was adequate to elicit changes in rDNA transcription and subsequently ribosome biogenesis by proxy. Further, the training, testing and biological sampling were conducted under standardized conditions, and there were no differences between pairwise consecutive days in macronutrient ingestion and training volume. As such, the day-to-day and test-to-test variation should have been minimal, especially when eliminating between-subjects biological variation with a within-subjects design as well (SOURCE). Importantly, both Mariappan (11) and Tanaka (15) compared high glucose treatment to either low glucose treatment or glucose starvation. In the present study however, we compared RT with either protein and glucose or protein and placebo. The participants fasted from 2200hrs the preceding day until the morning of each training day, receiving only the prescribed treatment until 2.5 hrs after RT. Thus, except from approximately 0700hrs-1230hrs, the participants ingested an equal amount of carbohydrates on pairwise days. As such, there were no differences in baseline levels of plasma glucose prior to the daily intervention, as by design, since we were interested in investigating if there were direct effects of glucose as a signalling molecule, based on the observations of Tanaka et al. [@tanaka_mild_2015] and Mariappan et al. [@mariappan_ribosomal_2011]. Further, the significant difference in levels of plasma glucose and c-peptide measured before, during and after session six, confirms that the treatment during of RT with glucose indeed was sufficient. Despite of this, we found no difference in markers of ribosome biogenesis between RT with glucose and RT with placebo, this does suggest that there are no additional effects of exercising with glucose, when total macronutrient intake is standardized.

Prior to the present study, it was theorized that glucose may convey an anabolic signal, either directly or through insulin, and thus operate as a signal mediator affecting upstream pathways of ribosome biogenesis [@kusnadi_regulation_2015; @sebata_control_2018]. This theoretical anabolic signal could possibly stimulate PIC assembly and Pol I recruitment to the rDNA promoter, initiating rDNA transcription and subsequent rRNA maturation, and lastly increased ribosome content of the skeletal muscle cell [@figueiredo_regulation_2019; @sebata_control_2018]. Thus, we hypothesized that if glucose directly or indirectly via insulin can stimulate initiation of rDNA transcription, five RT sessions with glucose will elicit this effect compared to five RT sessions with placebo.  This time frame has previously shown to be sufficient to measure changes in markers of ribosome biogenesis [@hammarstrom_benefits_2020; @hammarstrom_ribosome_2022; @figueiredo_impact_2016]. <!--double-check these references--> As no such effect was observable in the present study, this hypothesis was rejected. Further, glucose ingestion affects cellular energy levels; thus another theory is that high glucose treatment would stimulate energy-sensitive pathways such as PIH1, mTORC1, ERK1/2, AMPK and SIRT1 in a dose-response relationship [@kim_nutrient_2013; @mariappan_ribosomal_2011; @tanaka_mild_2015; @zhai_human_2012]. <!-- check if this is right --> Mammalian target of rapamycin complex 1, along with ERK1/2 and PIH1 upregulates PIC assembly during high-energy conditions, while AMPK and possibly SIRT1 represses assembly of the PIC during low-energy conditions, as observed in rodents [SOURCE], yeast [SOURCE], and human non-muscle cells [SOURCE] *in vitro*. Specifically, high glucose treatment augmented rDNA transcription by increased UBF and SL-1 activity, while low glucose treatment, glucose starvation, or nutrient depletion stimulated AMPK-dependent TIF-1A and chromatin regulation suppression [@kos-braun_tor1_2017: @mariappan_ribosomal_2011; @zhai_human_2012]. <!--check these references--> In addition to UBF and SL-1, assembly of the PIC requires TIF-1A and Pol I recruitment, induced by amino acid intake in an mTOR-S6K-dependent manner and possibly by direct c-Myc activation [@kusnadi_regulation_2015]. As such, increasing energy supply by RT with glucose vs. RT with placebo and nutrient availability should in line with these observations potently stimulate PIC assembly and rDNA transcription. Though we did not investigate these pathways directly, the downstream targets UBF and rpS6 did not differ in mean change between treatments. As the participants in the present study fasted until the daily training was completed, and matched their daily energy intake on pairwise days, this suggests that as long as the daily energy requirements are met, ingesting glucose with RT does not stimulate further gains in anabolic responses. Notably, the daily RT protocol was quite short (~30min), which may have reduced the effectiveness of the energy addition through glucose. In this regard, it would have been interesting to measure the intramuscular glycogen storages, to see how the RT sessions impacted intracellular energy levels. Unfortunately, such analyses were not conducted in the present study. Regrettably, this is a weakness of the design, as we have not measured directly what happened to the excess plasma glucose. Nevertheless, the absent effect of increasing energy availability during RT suggest that, there is no dose-response relationship between acutely increasing myocellular energy levels before and after RT and ribosome biogenesis.

Interestingly, we did observe a linear relationship between total RNA and UBF, suggesting ribosome biogenesis to be dependent on UBF activation, in line with observations in Mariappan et al. (2011). This was not evident with total RNA and c-Myc or rpS6, which we observed no linear relationship between. The upstream binding factor (UBF) has indeed previously been described as a master regulator of rDNA transcription, based on findings from cell cultures [@figueiredo_regulation_2019; @kusnadi_regulation_2015; @russell_rna-polymerase-i-directed_2005]. However, it has only recently been found in a linear relationship with total RNA content in human skeletal muscle cells after a short period of RT [@hammarstrom_ribosome_2022], which the observation of the present study supports. As such, UBF also seem to respond to mechanical loading in human muscle cells in line with responses seen in cell cultures and synergist ablation models [@mariappan_ribosomal_2011; @vonwalden18], in addition to the previously reported response to hormonal, nutritional, and cellular energy signals [@figueiredo_regulation_2019; @kusnadi_regulation_2015; @russell_rna-polymerase-i-directed_2005]. While c-Myc and rpS6 did not increase linearly with accumulation of total RNA, the cellular content of c-Myc and rpS6 increased from baseline to post with both treatments. The transcription initiation factor c-Myc has been implicated as important to ribosome biogenesis [@von_walden_ribosome_2019; @figueiredo_regulation_2019], though its role is likely as a general transcription factor rather than specific, compared to UBF [SOURCES?]. Further, the ribosomal protein S6 (rpS6) has previously been shown to correlate strongly with rRNA content after synergist ablation [@nakada_correlation_2016; @chaillou_hypoxia_2012], <!--check Nakada 16 for these refs--> and is suggested as a reliable measurement of ribosome biogenesis [@nakada_correlation_2016]. Together, this may explain why the between treatment difference in c-Myc content did not induce differences in total RNA or rRNA content, as the specific transcription factor UBF and the rpS6 increased equally with both treatments. Notably, the difference between c-Myc in glucose and placebo may have caused responses in pathways not measured in the present study.

The results from the strength tests do indicate that the participants might have been fatigued during the intervention, as they exercised every day for 12 days straight, with strength testing in-between. In this perspective, we expected glucose to increase energy availability and thus, muscle performance during strength testing compared to placebo.  However, this was not evident, as the mean response in strength as an index of peak torque at all speeds decreased with both treatments from baseline to post five and six sessions. Notably, RT with glucose reduced strength significantly less than RT with placebo post five sessions, which may indeed have been an effect of increased energy availability, seeing as glucose is the preferred energy source during strenuous activity [@mul_exercise_2015]. <!--check this ref--> Arguably, this might imply that RT with placebo experienced a higher stress response to the RT program, thus reduced muscle performance in strength tests. This could also explain the differences between treatments in content of the stress responsive c-Myc [SOURCES], meaning higher stress exerted during RT with placebo compared to RT with glucose inducing higher accumulation of c-Myc. If this was the case, it does seem inconclusive that strength did not differ between treatments at earlier or later time points. In addition, there were no differences in training volume per session between treatments, therefore, the mechanical loading should have been equal. Another potential explanation to this observation might be …. (bifasisk restitusjon?). Lastly, these inconclusive observations may also reflect our limited sample size, despite using a within-subjects design. Initially, we met our minimum limit of 16 participants however, three dropped out during the intervention, leaving us slightly underpowered according to the a priori calculation.

In conclusion, resistance exercise training combined with glucose ingestion does not acutely augment ribosome biogenesis following two weeks of heavy resistance training in moderately trained young male and female adults. Neither does it acutely enhance muscular performance or recovery withing 23hrs, measured as unilateral isometric and isokinetic knee extension peak torque. The observations in total RNA, rRNA and rDNA transcription initiation associated proteins confirm the key role of increased UBF content in ribosome biogenesis in moderately trained human skeletal muscle following resistance training, previously observed in RT-untrained participants, rodents or *in vitro* [SOURCES?]. If future investigations are to be made on this topic, a greater sample size coupled with a negative control group and analyses of the intramuscular glycogen storage may provide higher resolution results.

>>>>

Responses to systematic resistance training (RT) vary widely between individuals, with as many as 10-15% showing impaired muscle growth in response to standardized training interventions [@ALVAREZ18; @MANN14; @THALACKER13]. Such inability to benefit from RT may be explained by genetic and/or epigenetic disposition, affecting the internal physiological milieu in terms of ability to mount an anabolic response to RT [@figueiredo_genetic_2021;@thalacker-mercer_cluster_2013].<!-- Check these refs, do they correspond to genet epigen and internal mileu? Kristian: The way I read it, it looks like it corresponds --> Although RT program modifications such as altered exercise volume or intensity may induce more favorable adaptations in some individuals, such benefits are not uniform between individuals [@hammarstrom_benefits_2020;@krieger_single_2009;@schoenfeld_dose-response_2017].<!-- Add ref for more rt program modifications. Kristian: Added krieger et al. 2009 and schoenfeld et al. 2017 --> For instance, although changes in training volume generally induces greater benefits of RT, not all individuals benefit from such modifications [@hammarstrom_benefits_2020]. Therefore, other means than modification of RT variables seems necessary to optimize individual responses.<!-- The above sentence might be a bit contradictory? Im not sure, are you thinking about "other means than modification of RT" being contradictory of "modifications such as altered exercise volume.." sentence? I think there is a point to establishing that yes, increasing exercise volume is a potent way to increase the RT stimulus in general. However, in cases where this alone fails to elicit true responses, other means are necessary. Or am I missinterpreting this? -->

@@

To our knowledge, no previous study has investigated the effects of glucose compared to placebo ingestion with RT on ribosome biogenesis in human skeletal muscle, making our hypotheses exploratory and comparisons not straight forwards. Mariappan and colleagues [@mariappan_ribosomal_2011] investigated the effect of high glucose treatment on ribosome biogenesis in glomerular epithelial cells (GECs) from diabetic (type 1 and 2) mice, observing a UBF-dependent augmentation of ribosome biogenesis. In a study from 2015, Tanaka et al. [@tanaka_mild_2015] observed that mild glucose starvation reduced rRNA transcription in human breast cancer cells. These findings do imply an important role for glucose in the initiation of rDNA transcription, the rate-limiting step of ribosome biogenesis [@mariappan_ribosomal_2011; @tanaka_mild_2015]. In the present study, we did not find such indications of glucose having an effect as these previous studies. Notably, kidney cells from mice and human breast cancer cells are likely to have a higher physiological rate compared to human skeletal muscle cells, thus, our two-week intervention may not be directly comparable to these findings, as was expected. Nevertheless, we did observe robust increases in both total RNA and 47S pre-rRNA and all four mature rRNAs (18S, 28S, 5.8S, 5S) after five RT sessions with both treatments. This means that the time frame and training volume chosen for this design was adequate to elicit changes in rDNA transcription and subsequently ribosome biogenesis by proxy, as previously reported [@figueiredo_impact_2016; @hammarstrom_benefits_2020]. Further, the training, testing and biological sampling were conducted under standardized conditions, and there were no differences between pairwise consecutive days in macronutrient ingestion and training volume. As such, the day-to-day and test-to-test variation should have been minimal, especially when eliminating between-subjects biological variation with a within-subjects design as well <!--FIND REF-->. Importantly, both Mariappan [@mariappan_ribosomal_2011] and Tanaka [@tanaka_mild_2015] compared high glucose treatment to either low glucose treatment or glucose starvation. In the present study however, we compared RT with either protein and glucose or protein and placebo. The participants fasted from 2200hrs the preceding day until the morning of each training day, receiving only the prescribed treatment until 2.5 hrs after RT. Thus, except from approximately 0700hrs-1230hrs, the participants ingested an equal amount of carbohydrates on pairwise days. As such, there were no differences in baseline levels of plasma glucose prior to the daily intervention, as by design, since we were interested in investigating if there were direct effects of glucose as a signalling molecule, based on the observations of Tanaka et al. [@tanaka_mild_2015] and Mariappan et al. [@mariappan_ribosomal_2011]. Further, the significant difference in levels of plasma glucose and c-peptide measured before, during and after session six, confirms that the treatment during of RT with glucose indeed was sufficient. Despite of this, we found no difference in markers of ribosome biogenesis between RT with glucose and RT with placebo, this does suggest that there are no additional effects of exercising with glucose, when total macronutrient intake is standardized.

Prior to the present study, it was theorized that glucose may convey an anabolic signal, either directly or through insulin, and thus operate as a signal mediator affecting upstream pathways of ribosome biogenesis [@kusnadi_regulation_2015; @sebata_control_2018]. This theoretical anabolic signal could possibly stimulate PIC assembly and Pol I recruitment to the rDNA promoter, initiating rDNA transcription and subsequent rRNA maturation, and lastly increased ribosome content of the skeletal muscle cell [@figueiredo_regulation_2019; @sebata_control_2018]. Thus, we hypothesized that if glucose directly or indirectly via insulin can stimulate initiation of rDNA transcription, five RT sessions with glucose will elicit this effect compared to five RT sessions with placebo.  This time frame has previously shown to be sufficient to measure changes in markers of ribosome biogenesis [@hammarstrom_benefits_2020; @hammarstrom_ribosome_2022; @figueiredo_impact_2016]. <!--double-check these references--> As no such effect was observable in the present study, this hypothesis was rejected. Further, glucose ingestion affects cellular energy levels; thus another theory is that high glucose treatment would stimulate energy-sensitive pathways such as PIH1, mTORC1, ERK1/2, AMPK and SIRT1 in a dose-response relationship [@kim_nutrient_2013; @mariappan_ribosomal_2011; @tanaka_mild_2015; @zhai_human_2012]. <!-- check if this is right --> Mammalian target of rapamycin complex 1, along with ERK1/2 and PIH1 upregulates PIC assembly during high-energy conditions, while AMPK and possibly SIRT1 represses assembly of the PIC during low-energy conditions, as observed in rodents <!-- REF -->, yeast <!-- REF -->, and human non-muscle cells *in vitro*. <!-- REF --> Specifically, high glucose treatment augmented rDNA transcription by increased UBF and SL-1 activity, while low glucose treatment, glucose starvation, or nutrient depletion stimulated AMPK-dependent TIF-1A and chromatin regulation suppression [@kos-braun_tor1_2017: @mariappan_ribosomal_2011; @zhai_human_2012]. <!--check these references--> In addition to UBF and SL-1, assembly of the PIC requires TIF-1A and Pol I recruitment, induced by amino acid intake in an mTOR-S6K-dependent manner and possibly by direct c-Myc activation [@kusnadi_regulation_2015]. As such, increasing energy supply by RT with glucose vs. RT with placebo and nutrient availability should in line with these observations potently stimulate PIC assembly and rDNA transcription. Though we did not investigate these pathways directly, the downstream targets UBF and rpS6 did not differ in mean change between treatments. As the participants in the present study fasted until the daily training was completed, and matched their daily energy intake on pairwise days, this suggests that as long as the daily energy requirements are met, ingesting glucose with RT does not stimulate further gains in anabolic responses. Notably, the daily RT protocol was quite short (~30min), which may have reduced the effectiveness of the energy addition through glucose. In this regard, it would have been interesting to measure the intramuscular glycogen storages, to see how the RT sessions impacted intracellular energy levels. Unfortunately, such analyses were not conducted in the present study. Regrettably, this is a weakness of the design, as we have not measured directly what happened to the excess plasma glucose. Nevertheless, the absent effect of increasing energy availability during RT suggest that, there is no dose-response relationship between acutely increasing myocellular energy levels before and after RT and ribosome biogenesis.

Interestingly, we did observe a linear relationship between total RNA and UBF, suggesting ribosome biogenesis to be dependent on UBF activation, in line with observations in Mariappan et al. (2011). This was not evident with total RNA and c-Myc or rpS6, which we observed no linear relationship between. The upstream binding factor (UBF) has indeed previously been described as a master regulator of rDNA transcription, based on findings from cell cultures [@figueiredo_regulation_2019; @kusnadi_regulation_2015; @russell_rna-polymerase-i-directed_2005]. However, it has only recently been found in a linear relationship with total RNA content in human skeletal muscle cells after a short period of RT [@hammarstrom_ribosome_2022], which the observation of the present study supports. As such, UBF also seem to respond to mechanical loading in human muscle cells in line with responses seen in cell cultures and synergist ablation models [@mariappan_ribosomal_2011; @von_walden_mechanical_2012], in addition to the previously reported response to hormonal, nutritional, and cellular energy signals [@figueiredo_regulation_2019; @kusnadi_regulation_2015; @russell_rna-polymerase-i-directed_2005]. While c-Myc and rpS6 did not increase linearly with accumulation of total RNA, the cellular content of c-Myc and rpS6 increased from baseline to post with both treatments. The transcription initiation factor c-Myc has been implicated as important to ribosome biogenesis [@von_walden_ribosome_2019; @figueiredo_regulation_2019], though its role is likely as a general transcription factor rather than specific, compared to UBF<!-- REF -->. Further, the ribosomal protein S6 (rpS6) has previously been shown to correlate strongly with rRNA content after synergist ablation [@nakada_correlation_2016; @chaillou_hypoxia_2012], and is suggested as a reliable measurement of ribosome biogenesis [@nakada_correlation_2016]. Together, this may explain why the between treatment difference in c-Myc content did not induce differences in total RNA or rRNA content, as the specific transcription factor UBF and the rpS6 increased equally with both treatments. Notably, the difference between c-Myc in glucose and placebo may have caused responses in pathways not measured in the present study.

The results from the strength tests do indicate that the participants might have been fatigued during the intervention, as they exercised every day for 12 days straight, with strength testing in-between. In this perspective, we expected glucose to increase energy availability and thus, muscle performance during strength testing compared to placebo. However, this was not evident, as the mean response in strength as an index of peak torque at all speeds decreased with both treatments from baseline to post five and six sessions. Notably, RT with glucose reduced strength significantly less than RT with placebo post five sessions, which may indeed have been an effect of increased energy availability, seeing as glucose is the preferred energy source during strenuous activity [@mul_exercise_2015]. <!--check this ref--> Arguably, this might imply that RT with placebo experienced a higher stress response to the RT program, thus reduced muscle performance in strength tests. This could also explain the differences between treatments in content of the stress responsive c-Myc <!-- REF -->, meaning higher stress exerted during RT with placebo compared to RT with glucose inducing higher accumulation of c-Myc. If this was the case, it does seem inconclusive that strength did not differ between treatments at earlier or later time points. In addition, there were no differences in training volume per session between treatments, therefore, the mechanical loading should have been equal. Another potential explanation to this observation might be …. (bifasisk restitusjon?)<!-- investigate this -->. Lastly, these inconclusive observations may also reflect our limited sample size, despite using a within-subjects design. Initially, we met our minimum limit of 16 participants however, three dropped out during the intervention, leaving us slightly underpowered according to the a priori calculation.

In conclusion, resistance exercise training combined with glucose ingestion does not acutely augment ribosome biogenesis following two weeks of heavy resistance training in moderately trained young male and female adults. Neither does it acutely enhance muscular performance or recovery withing 23hrs, measured as unilateral isometric and isokinetic knee extension peak torque. The observations in total RNA, rRNA and rDNA transcription initiation associated proteins confirm the key role of increased UBF content in ribosome biogenesis in moderately trained human skeletal muscle following resistance training, previously observed in RT-untrained participants, rodents or *in vitro* <!-- REF -->. If future investigations are to be made on this topic, a greater sample size coupled with a negative control group and analyses of the intramuscular glycogen storage may provide higher resolution results.
